---
title: "Hands-on Exercise 4"
author: "Leng Yi Xiu"
date: "Feb 02 2025"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  message: false
  freeze: true
---

## 1 Visualising Distribution

### 1.1 Learning Outcome

Visualising distribution is not new in statistical analysis. In chapter 1 we have shared with you some of the popular statistical graphics methods for visualising distribution are histogram, probability density curve (pdf), boxplot, notch plot and violin plot and how they can be created by using ggplot2. In this chapter, we are going to share with you two relatively new statistical graphic methods for visualising distribution, namely ridgeline plot and raincloud plot by using ggplot2 and its extensions.

### 1.2 Loading R packages and data import

Loading R packages:

```{r}
pacman::p_load(ggdist, ggridges, ggthemes,
               colorspace, tidyverse)
```

Importing Data:

```{r}
exam <- read_csv("data/Exam_data.csv")
```

### 1.3 Visualising Distribution with Ridgeline Plot

[**Ridgeline plot**](https://www.data-to-viz.com/graph/ridgeline.html) (sometimes called *Joyplot*) is a data visualisation technique for revealing the distribution of a numeric value for several groups. Distribution can be represented using histograms or density plots, all aligned to the same horizontal scale and presented with a slight overlap.

::: {.callout-note appearance="simple"}
-   Ridgeline plots make sense when the number of group to represent is **medium to high**, and thus a classic window separation would take to much space. Indeed, the fact that groups overlap each other allows to use space more efficiently. If you have less than 5 groups, dealing with other distribution plots is probably better.

-   It works well when there is a clear pattern in the result, like if there is an obvious ranking in groups. Otherwise group will tend to overlap each other, leading to a messy plot not providing any insight.
:::

#### 1.3.1 Plotting ridgeline graph: ggridges method

ggridges package provides two main geom to plot gridgeline plots, they are: [`geom_ridgeline()`](https://wilkelab.org/ggridges/reference/geom_ridgeline.html) and [`geom_density_ridges()`](https://wilkelab.org/ggridges/reference/geom_density_ridges.html). The former takes height values directly to draw the ridgelines, and the latter first estimates data densities and then draws those using ridgelines.

Using `geom_density_ridges()`:

```{r}
ggplot(exam, 
       aes(x = ENGLISH, 
           y = CLASS)) +
  geom_density_ridges(
    scale = 3,
    rel_min_height = 0.01,
    bandwidth = 3.4,
    fill = lighten("#7097BB", .3),
    color = "white"
  ) +
  scale_x_continuous(
    name = "English grades",
    expand = c(0, 0)
    ) +
  scale_y_discrete(name = NULL, expand = expansion(add = c(0.2, 2.6))) +
  theme_ridges()
```

#### 1.3.2 Varying fill colors along the x axis

```{r}
ggplot(exam, 
       aes(x = ENGLISH, 
           y = CLASS,
           fill = stat(x))) +
  geom_density_ridges_gradient(
    scale = 3,
    rel_min_height = 0.01) +
  scale_fill_viridis_c(name = "Temp. [F]",
                       option = "C") +
  scale_x_continuous(
    name = "English grades",
    expand = c(0, 0)
  ) +
  scale_y_discrete(name = NULL, expand = expansion(add = c(0.2, 2.6))) +
  theme_ridges()
```

#### 1.3.3 Mapping the probabilities directly onto colour

```{r}
ggplot(exam,
       aes(x = ENGLISH, 
           y = CLASS, 
           fill = 0.5 - abs(0.5-stat(ecdf)))) +
  stat_density_ridges(geom = "density_ridges_gradient", 
                      calc_ecdf = TRUE) +
  scale_fill_viridis_c(name = "Tail probability",
                       direction = -1) +
  theme_ridges()
```

::: {.callout-caution appearance="simple"}
It is important include the argument `calc_ecdf = TRUE` in `stat_density_ridges()`.
:::

#### 1.3.4 Ridgeline plots with quantile lines

```{r}
ggplot(exam,
       aes(x = ENGLISH, 
           y = CLASS, 
           fill = factor(stat(quantile))
           )) +
  stat_density_ridges(
    geom = "density_ridges_gradient",
    calc_ecdf = TRUE, 
    quantiles = 4,
    quantile_lines = TRUE) +
  scale_fill_viridis_d(name = "Quartiles") +
  theme_ridges()
```

Using cut points such as 2.5% and 97.5% tails to colour the ridgeline plot:

```{r}
ggplot(exam,
       aes(x = ENGLISH, 
           y = CLASS, 
           fill = factor(stat(quantile))
           )) +
  stat_density_ridges(
    geom = "density_ridges_gradient",
    calc_ecdf = TRUE, 
    quantiles = c(0.025, 0.975)
    ) +
  scale_fill_manual(
    name = "Probability",
    values = c("#FF0000A0", "#A0A0A0A0", "#0000FFA0"),
    labels = c("(0, 0.025]", "(0.025, 0.975]", "(0.975, 1]")
  ) +
  theme_ridges()
```

### 1.4 Visualising Distribution with Raincloud Plot

Raincloud Plot is a data visualisation techniques that produces a half-density to a distribution plot. It gets the name because the density plot is in the shape of a “raincloud”. The raincloud (half-density) plot enhances the traditional box-plot by highlighting multiple modalities (an indicator that groups may exist). The boxplot does not show where densities are clustered, but the raincloud plot does!

#### 1.4.1 Plotting a Half Eye graph

```{r}
ggplot(exam, 
       aes(x = RACE, 
           y = ENGLISH)) +
  stat_halfeye(adjust = 0.5,
               justification = -0.2,
               .width = 0,
               point_colour = NA)
```

#### 1.4.2 Adding the boxplot with `geom_boxplot()`

```{r}
ggplot(exam, 
       aes(x = RACE, 
           y = ENGLISH)) +
  stat_halfeye(adjust = 0.5,
               justification = -0.2,
               .width = 0,
               point_colour = NA) +
  geom_boxplot(width = .20,
               outlier.shape = NA)
```

#### 1.4.3 Adding the Dot Plots with `stat_dots()`

```{r}
ggplot(exam, 
       aes(x = RACE, 
           y = ENGLISH)) +
  stat_halfeye(adjust = 0.5,
               justification = -0.2,
               .width = 0,
               point_colour = NA) +
  geom_boxplot(width = .20,
               outlier.shape = NA) +
  stat_dots(side = "left", 
            justification = 1.2, 
            binwidth = .5,
            dotsize = 2)
```

#### 1.4.4 Finishing touch

```{r}
ggplot(exam, 
       aes(x = RACE, 
           y = ENGLISH)) +
  stat_halfeye(adjust = 0.5,
               justification = -0.2,
               .width = 0,
               point_colour = NA) +
  geom_boxplot(width = .20,
               outlier.shape = NA) +
  stat_dots(side = "left", 
            justification = 1.2, 
            binwidth = .5,
            dotsize = 1.5) +
  coord_flip() +
  theme_economist()
```

## 2 Visual Statistical Analysis

### 2.1 Learning Outcome

-   ggstatsplot package to create visual graphics with rich statistical information,

-   performance package to visualise model diagnostics, and

-   parameters package to visualise model parameters

### 2.2 Visual Statistical Analysis with ggstatsplot

**ggstatsplot** is an extension of [**ggplot2**](https://ggplot2.tidyverse.org/) package for creating graphics with details from statistical tests included in the information-rich plots themselves.

### 2.3 Getting Started

#### 2.3.1 Loading R package

```{r}
pacman::p_load(ggstatsplot, tidyverse)
```

#### 2.3.2 Import Data

```{r}
exam <- read_csv("data/Exam_data.csv")
exam
```

#### 2.3.3 One-sample test: `gghistostats()` method

```{r}
set.seed(1234)

gghistostats(
  data = exam,
  x = ENGLISH,
  type = "bayes",
  test.value = 60,
  xlab = "English scores"
)
```

#### 2.3.6 Two-sample mean test: `ggbetweenstats()` method

```{r}
ggbetweenstats(
  data = exam,
  x = GENDER, 
  y = MATHS,
  type = "np",
  messages = FALSE
)
```

#### 2.3.7 Oneway ANOVA test: `ggbetweenstats()` method

```{r}
ggbetweenstats(
  data = exam,
  x = RACE, 
  y = ENGLISH,
  type = "p",
  mean.ci = TRUE, 
  pairwise.comparisons = TRUE, 
  pairwise.display = "s",
  p.adjust.method = "fdr",
  messages = FALSE
)
```

#### 2.3.8 Significant test of Correlation: `ggscatterstats()`

```{r}
ggscatterstats(
  data = exam,
  x = MATHS,
  y = ENGLISH,
  marginal = FALSE,
  )
```

#### 2.3.9 Significant test of Association (Dependence): `ggbarstats()` methods

Maths scores is binned into a 4-class variable:

```{r}
exam1 <- exam %>% 
  mutate(MATHS_bins = 
           cut(MATHS, 
               breaks = c(0,60,75,85,100))
)
```

Visual for Significant Test of Association:

```{r}
ggbarstats(exam1, 
           x = MATHS_bins, 
           y = GENDER)
```

### 2.4 Visualising Models

-   Visualising model diagnostic and model parameters by using parameters package

### 2.5 Loading R packages and data import

```{r}
pacman::p_load(readxl, performance, parameters, see)
```

```{r}
car_resale <- read_xls("data/ToyotaCorolla.xls", 
                       "data")
car_resale
```

### 2.6 Model Diagnostic

#### 2.6.1 Multiple Regression Model using `lm()`

```{r}
model <- lm(Price ~ Age_08_04 + Mfg_Year + KM + 
              Weight + Guarantee_Period, data = car_resale)
model
```

#### 2.6.2 Checking for multicollinearity

```{r}
check_c <- check_collinearity(model)
check_c
plot(check_c)
```

#### 2.6.3 Checking for normality assumption

```{r}
model1 <- lm(Price ~ Age_08_04 + KM + 
              Weight + Guarantee_Period, data = car_resale)

check_n <- check_normality(model1)

plot(check_n)
```

#### 2.6.4 Checking for homogeneity of variances

```{r}
check_h <- check_heteroscedasticity(model1)

plot(check_h)
```

#### 2.6.5 Complete check

```{r}
check_model(model1)
```

#### 2.6.6 Visualising Regression Parameters: see methods

```{r}
plot(parameters(model1))
```

#### 2.6.7 Visualising Regression Parameters: `ggcoefstats()` methods

```{r}
ggcoefstats(model1, 
            output = "plot")
```

## 3 Visualising Uncertainty

### 3.1 Learning Outcome

-   Plot statistics error bars by using **ggplot2**,

-   Plot interactive error bars by combining **ggplot2**, **plotly** and **DT**,

-   Create advanced by using **ggdist**, and

-   Create hypothetical outcome plots (HOPs) by using **ungeviz** package.

### 3.2 Loading R Packages and Import Data

```{r}
pacman::p_load(plotly, crosstalk, DT, 
               ggdist, ggridges, colorspace,
               gganimate, tidyverse)
```

```{r}
exam <- read_csv("data/Exam_data.csv")
```

### 3.3 Visualising the uncertainty of point estimates: ggplot2 methods

A point estimate is a single number, such as a mean. Uncertainty, on the other hand, is expressed as standard error, confidence interval, or credible interval.

```{r}
my_sum <- exam %>%
  group_by(RACE) %>%
  summarise(
    n=n(),
    mean=mean(MATHS),
    sd=sd(MATHS)
    ) %>%
  mutate(se=sd/sqrt(n-1))
```

::: {.callout-tip appearance="simple"}
-   `group_by()` of **dplyr** package is used to group the observation by RACE,

-   `summarise()` is used to compute the count of observations, mean, standard deviation

-   `mutate()` is used to derive standard error of Maths by RACE, and

-   the output is save as a tibble data table called *my_sum*.
:::

Display tibble data frame in html table format:

```{r}
knitr::kable(head(my_sum), format = 'html')
```

#### 3.3.1 Plotting standard error bars of point estimates

```{r}
ggplot(my_sum) +
  geom_errorbar(
    aes(x=RACE, 
        ymin=mean-se, 
        ymax=mean+se), 
    width=0.2, 
    colour="black", 
    alpha=0.9, 
    linewidth=0.5) +
  geom_point(aes
           (x=RACE, 
            y=mean), 
           stat="identity", 
           color="red",
           size = 1.5,
           alpha=1) +
  ggtitle("Standard error of mean maths score by rac")
```

#### 3.3.2 Plotting confidence interval of point estimates

```{r}
ggplot(my_sum) +
  geom_errorbar(
    aes(x=reorder(RACE, -mean), 
        ymin=mean-1.96*se, 
        ymax=mean+1.96*se), 
    width=0.2, 
    colour="black", 
    alpha=0.9, 
    linewidth=0.5) +
  geom_point(aes
           (x=RACE, 
            y=mean), 
           stat="identity", 
           color="red",
           size = 1.5,
           alpha=1) +
  labs(x = "Maths score",
       title = "95% confidence interval of mean maths score by race")
```

#### 3.3.3 Visualising the uncertainty of point estimates with interactive error bars

```{r}
shared_df = SharedData$new(my_sum)

bscols(widths = c(4,8),
       ggplotly((ggplot(shared_df) +
                   geom_errorbar(aes(
                     x=reorder(RACE, -mean),
                     ymin=mean-2.58*se, 
                     ymax=mean+2.58*se), 
                     width=0.2, 
                     colour="black", 
                     alpha=0.9, 
                     size=0.5) +
                   geom_point(aes(
                     x=RACE, 
                     y=mean, 
                     text = paste("Race:", `RACE`, 
                                  "<br>N:", `n`,
                                  "<br>Avg. Scores:", round(mean, digits = 2),
                                  "<br>95% CI:[", 
                                  round((mean-2.58*se), digits = 2), ",",
                                  round((mean+2.58*se), digits = 2),"]")),
                     stat="identity", 
                     color="red", 
                     size = 1.5, 
                     alpha=1) + 
                   xlab("Race") + 
                   ylab("Average Scores") + 
                   theme_minimal() + 
                   theme(axis.text.x = element_text(
                     angle = 45, vjust = 0.5, hjust=1)) +
                   ggtitle("99% Confidence interval of average /<br>maths scores by race")), 
                tooltip = "text"), 
       DT::datatable(shared_df, 
                     rownames = FALSE, 
                     class="compact", 
                     width="100%", 
                     options = list(pageLength = 10,
                                    scrollX=T), 
                     colnames = c("No. of pupils", 
                                  "Avg Scores",
                                  "Std Dev",
                                  "Std Error")) %>%
         formatRound(columns=c('mean', 'sd', 'se'),
                     digits=2))
```

### 3.4 Visualising Uncertainty: ggdist package

#### 3.4.1 Visualizing the uncertainty of point estimates: ggdist methods

```{r}
exam %>%
  ggplot(aes(x = RACE, y = MATHS)) +
  stat_pointinterval(.width = 0.95,
  .point = median,
  .interval = qi) +
  labs(
    title = "Visualising confidence intervals of median math score",
    subtitle = "Median Point + Multiple-interval plot")
```

#### 3.4.2 Visualizing the uncertainty of point estimates: ggdist methods

::: {.callout-note appearance="simple"}
Makeover the plot on previous slide by showing 95% and 99% confidence intervals:
:::

```{r}
exam %>%
  ggplot(aes(x = RACE, y = MATHS)) +
  stat_pointinterval(
    .width = c(0.95, 0.99), 
    .point = median,
    .interval = qi
  ) +
  labs(
    title = "Visualising confidence intervals of median math score",
    subtitle = "Median Point + Multiple-interval plot"
  )

```

#### 3.4.3 Visualizing the uncertainty of point estimates: ggdist methods

Visual for displaying distribution of maths scores by race:

```{r}
exam %>%
  ggplot(aes(x = RACE, 
             y = MATHS)) +
  stat_gradientinterval(   
    fill = "skyblue",      
    show.legend = TRUE     
  ) +                        
  labs(
    title = "Visualising confidence intervals of mean math score",
    subtitle = "Gradient + interval plot")
```

### 3.5 Visualising Uncertainty with Hypothetical Outcome Plots (HOPs)

#### 3.5.1 Installing ungeviz package

```{r}
devtools::install_github("wilkelab/ungeviz")
```

#### 3.5.2 Launch the application in R

```{r}
library(ungeviz)
```

#### 3.5.3 Visualising Uncertainty with Hypothetical Outcome Plots (HOPs)

```{r}
ggplot(data = exam, 
       (aes(x = factor(RACE), 
            y = MATHS))) +
  geom_point(position = position_jitter(
    height = 0.3, 
    width = 0.05), 
    size = 0.4, 
    color = "#0072B2", 
    alpha = 1/2) +
  geom_hpline(data = sampler(25, 
                             group = RACE), 
              height = 0.6, 
              color = "#D55E00") +
  theme_bw() + 
  transition_states(.draw, 1, 3)
```

## 4 Funnel Plots for Fair Comparisons

### 4.1 Learning Outcomes

-   Plotting funnel plots by using **funnelPlotR** package,

-   Plotting static funnel plot by using ggplot2 package, and

-   Plotting interactive funnel plot by using both **plotly R** and **ggplot2** packages

### 4.2 Loading R Packages

```{r}
pacman::p_load(tidyverse, FunnelPlotR, plotly, knitr)
```

### 4.3 Import Data

```{r}
covid19 <- read_csv("data/COVID-19_DKI_Jakarta.csv") %>%
  mutate_if(is.character, as.factor)
```

### 4.4 FunnelPlotR methods

[**FunnelPlotR**](https://nhs-r-community.github.io/FunnelPlotR/) package uses ggplot to generate funnel plots. It requires a `numerator` (events of interest), `denominator` (population to be considered) and `group`.

#### 4.4.1 FunnelPlotR methods: The basic plot

```{r}
funnel_plot(
  .data = covid19,
  numerator = Positive,
  denominator = Death,
  group = `Sub-district`
)
```

::: {.callout-note appearance="simple"}
-   `group` in this function is different from the scatterplot. Here, it defines the level of the points to be plotted i.e. Sub-district, District or City. If Cityc is chosen, there are only six data points.

-   By default, `data_type`argument is “SR”.

-   `limit`: Plot limits, accepted values are: 95 or 99, corresponding to 95% or 99.8% quantiles of the distribution.
:::

#### 4.4.2 FunnelPlotR methods: Makeover 1

```{r}
funnel_plot(
  .data = covid19,
  numerator = Death,
  denominator = Positive,
  group = `Sub-district`,
  data_type = "PR",     
  xrange = c(0, 6500),  
  yrange = c(0, 0.05)   
)
```

#### 4.4.3 FunnelPlotR methods: Makeover 2

```{r}
funnel_plot(
  .data = covid19,
  numerator = Death,
  denominator = Positive,
  group = `Sub-district`,
  data_type = "PR",   
  xrange = c(0, 6500),  
  yrange = c(0, 0.05),
  label = NA,
  title = "Cumulative COVID-19 Fatality Rate by Cumulative Total Number of COVID-19 Positive Cases",           
  x_label = "Cumulative COVID-19 Positive Cases", 
  y_label = "Cumulative Fatality Rate"
)
```

### 4.5 Funnel Plot for Fair Visual Comparison: ggplot2 methods

#### 4.5.1 Computing the basic derived fields

Derive cumulative death rate and standard error of cumulative death rate:

```{r}
df <- covid19 %>%
  mutate(rate = Death / Positive) %>%
  mutate(rate.se = sqrt((rate*(1-rate)) / (Positive))) %>%
  filter(rate > 0)
```

```{r}
fit.mean <- weighted.mean(df$rate, 1/df$rate.se^2)
```

#### 4.5.2 Calculate lower and upper limits for 95% and 99.9% CI

```{r}
number.seq <- seq(1, max(df$Positive), 1)
number.ll95 <- fit.mean - 1.96 * sqrt((fit.mean*(1-fit.mean)) / (number.seq)) 
number.ul95 <- fit.mean + 1.96 * sqrt((fit.mean*(1-fit.mean)) / (number.seq)) 
number.ll999 <- fit.mean - 3.29 * sqrt((fit.mean*(1-fit.mean)) / (number.seq)) 
number.ul999 <- fit.mean + 3.29 * sqrt((fit.mean*(1-fit.mean)) / (number.seq)) 
dfCI <- data.frame(number.ll95, number.ul95, number.ll999, 
                   number.ul999, number.seq, fit.mean)
```

#### 4.5.3 Plotting a static funnel plot

```{r}
p <- ggplot(df, aes(x = Positive, y = rate)) +
  geom_point(aes(label=`Sub-district`), 
             alpha=0.4) +
  geom_line(data = dfCI, 
            aes(x = number.seq, 
                y = number.ll95), 
            size = 0.4, 
            colour = "grey40", 
            linetype = "dashed") +
  geom_line(data = dfCI, 
            aes(x = number.seq, 
                y = number.ul95), 
            size = 0.4, 
            colour = "grey40", 
            linetype = "dashed") +
  geom_line(data = dfCI, 
            aes(x = number.seq, 
                y = number.ll999), 
            size = 0.4, 
            colour = "grey40") +
  geom_line(data = dfCI, 
            aes(x = number.seq, 
                y = number.ul999), 
            size = 0.4, 
            colour = "grey40") +
  geom_hline(data = dfCI, 
             aes(yintercept = fit.mean), 
             size = 0.4, 
             colour = "grey40") +
  coord_cartesian(ylim=c(0,0.05)) +
  annotate("text", x = 1, y = -0.13, label = "95%", size = 3, colour = "grey40") + 
  annotate("text", x = 4.5, y = -0.18, label = "99%", size = 3, colour = "grey40") + 
  ggtitle("Cumulative Fatality Rate by Cumulative Number of COVID-19 Cases") +
  xlab("Cumulative Number of COVID-19 Cases") + 
  ylab("Cumulative Fatality Rate") +
  theme_light() +
  theme(plot.title = element_text(size=12),
        legend.position = c(0.91,0.85), 
        legend.title = element_text(size=7),
        legend.text = element_text(size=7),
        legend.background = element_rect(colour = "grey60", linetype = "dotted"),
        legend.key.height = unit(0.3, "cm"))
p
```

#### 4.5.4 Interactive Funnel Plot: plotly + ggplot2

```{r}
fp_ggplotly <- ggplotly(p,
  tooltip = c("label", 
              "x", 
              "y"))
fp_ggplotly
```
