---
title: "Hands-on Exercise 3b"
author: "Leng Yi Xiu"
date: "Jan 30 2025"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  message: false
  freeze: true
---

## Programming Animated Statistical Graphics with R

### 4.1 Overview

When telling a visually-driven data story, animated graphics tends to attract the interest of the audience and make deeper impression than static graphics. In this hands-on exercise, you will learn how to create animated data visualisation by using **gganimate** and **plotly r** packages. At the same time, you will also learn how to (i) reshape data by using **tidyr** package, and (ii) process, wrangle and transform data by using **dplyr** package.

Before we dive into the steps for creating an animated statistical graph, it’s important to understand some of the key concepts and terminology related to this type of visualization.

1.  **Frame**: In an animated line graph, each frame represents a different point in time or a different category. When the frame changes, the data points on the graph are updated to reflect the new data.

2.  **Animation Attributes**: The animation attributes are the settings that control how the animation behaves. For example, you can specify the duration of each frame, the easing function used to transition between frames, and whether to start the animation from the current frame or from the beginning.

### 4.2 Getting Started

#### 4.2.1 Loading R Packages

```{r}
pacman::p_load(readxl, gifski, gapminder,
               plotly, gganimate, tidyverse)
```

#### 4.2.2 Importing the data

::: panel-tabset
## mutate_at()

```{r}
col <- c("Country", "Continent")
globalPop <- read_xls("data/GlobalPopulation.xls",
                      sheet="Data") %>%
  mutate_at(col, as.factor) %>%
  mutate(Year = as.integer(Year))
```

## across()

```{r}
col <- c("Country", "Continent")
globalPop <- read_xls("data/GlobalPopulation.xls",
                      sheet="Data") %>%
  mutate(across(col, as.factor)) %>%
  mutate(Year = as.integer(Year))
```
:::

### 4.3 Animated Data Visualisation: gganimate methods

[gganimate](https://gganimate.com/) extends the grammar of graphics as implemented by ggplot2 to include the description of animation. It does this by providing a range of new grammar classes that can be added to the plot object in order to customise how it should change with time.

#### 4.3.1 Building a static population bubble plot

```{r}
ggplot(globalPop, aes(x = Old, y = Young, 
                      size = Population, 
                      colour = Country)) +
  geom_point(alpha = 0.7, 
             show.legend = FALSE) +
  scale_colour_manual(values = country_colors) +
  scale_size(range = c(2, 12)) +
  labs(title = 'Year: {frame_time}', 
       x = '% Aged', 
       y = '% Young') 
```

#### 4.3.2 Building the animated bubble plot

```{r}
ggplot(globalPop, aes(x = Old, y = Young, 
                      size = Population, 
                      colour = Country)) +
  geom_point(alpha = 0.7, 
             show.legend = FALSE) +
  scale_colour_manual(values = country_colors) +
  scale_size(range = c(2, 12)) +
  labs(title = 'Year: {frame_time}', 
       x = '% Aged', 
       y = '% Young') +
  transition_time(Year) +       
  ease_aes('linear')          
```

### 4.4 Animated Data Visualisation: plotly

In **Plotly R** package, both `ggplotly()` and `plot_ly()` support key frame animations through the `frame` argument/aesthetic. They also support an `ids` argument/aesthetic to ensure smooth transitions between objects with the same id (which helps facilitate object constancy).

#### 4.4.1 Building the animated bubble plot: `ggplotly()` method

```{r}
gg <- ggplot(globalPop, 
       aes(x = Old, 
           y = Young, 
           size = Population, 
           colour = Country)) +
  geom_point(aes(size = Population,
                 frame = Year),
             alpha = 0.7) +
  scale_colour_manual(values = country_colors) +
  scale_size(range = c(2, 12)) +
  labs(x = '% Aged', 
       y = '% Young') + 
  theme(legend.position='none')

ggplotly(gg)
```

::: {.callout-important appearance="simple"}
`theme(legend.position='none')` used to hide the legend.

`show.legend = false` argument will not work in this instance.
:::

#### 4.4.2 Building the animated bubble plot: `plot_ly()` method

```{r}
bp <- globalPop %>%
  plot_ly(x = ~Old, 
          y = ~Young, 
          size = ~Population, 
          color = ~Continent,
          sizes = c(2, 100),
          frame = ~Year, 
          text = ~Country, 
          hoverinfo = "text",
          type = 'scatter',
          mode = 'markers'
          ) %>%
  layout(showlegend = FALSE)
bp
```
