---
title: "Take-home Exercise 3: Prototyping Modules for Visual Analytics Shiny Application"
author: "Leng Yi Xiu"
date: "Mar 22 2025"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  message: false
  freeze: true
---

## 1 Overview of Task

In this take-home exercise, we are required to select one of the module of our proposed Shiny application and complete the following tasks:

-   To evaluate and determine the necessary R packages needed for your Shiny application are supported in R CRAN,

-   To prepare and test the specific R codes can be run and returned the correct output as expected,

-   To determine the parameters and outputs that will be exposed on the Shiny applications, and

-   To select the appropriate Shiny UI components for exposing the parameters determine above.

1️⃣ This take-home exercise must provide detail discussion and explanation of:

-   the data preparation process,

-   the selection of data visualisation techniques used,

-   and the data visualisation design and interactivity principles and best practices implemented.

2️⃣ It should also include a section called **UI design** for the different components of the UIs for the proposed design.

Our group has chosen to develop an interactive R Shiny application (Weather Pulse) designed to enhance the exploration and understanding of **Singapore’s evolving climate** by addressing gaps in historical trend analysis, spatial insights, and future projections.

## 2 Data Preparation

### 2.1 The Dataset

The dataset from [MSS website](https://www.weather.gov.sg/climate-historical-daily/) features extensive historical daily records that include measurements of daily rainfall, temperature, and wind speed. These records are collected from 64 weather stations across Singapore, covering a period from 1980 to 2025.

As the data is stored disparately on the website, segmented by both month and by weather station, it must be scraped from the website and consolidated into a single dataset to be effectively utilised.

**R packages** that will be used for the data scraping and consolidation process are listed below:

1.  `dplyr`: This will be used in the data consolidation process as it offers functions for data manipulation and transformation.

2.  `lubridate`: This will be used for parsing, manipulating, and performing arithmetic on date-time objects.

```{r}
pacman::p_load(readxl, dplyr, readr, lubridate)
```

## 2.2 Scraping data from MSS website

The goal of our R Shiny application is to examine and analyze weather trends in Singapore **from 2020 to 2024**. To achieve this, we will first scrape the data readings of 20 automated weather stations across Singapore from the MSS website and then proceed with data wrangling.

```{r}
#| code-fold: true
#| code-summary: "Click to view code for reading list of stations"
stations_df <- read_excel("data/StationList.xlsx")  
station_codes <- stations_df$StationID   
```

```{r}
#| code-fold: true
#| code-summary: "Click to view code for data scraping"
# setting Year and Month
years <- 2020:2024
months <- sprintf("%02d", 1:12)

# Base URL Template
base_url <- "https://www.weather.gov.sg/files/dailydata/DAILYDATA_%s_%d%s.csv"

# Initialize an empty list
all_data <- list()
i <- 1

# Loop through 
for (station in station_codes) {
  for (year in years) {
    for (month in months) {
      url <- sprintf(base_url, station, year, month)
      cat("Downloading:", url, "\n")
    
    try({
      res <- read_csv(url, show_col_types = FALSE) %>%
        mutate(across(
          contains(c("Temperature", "Rainfall", "Wind")),
          ~ suppressWarnings(as.numeric(.))
        )) %>%
        mutate(
          station = as.character(station),
          year = as.integer(year),
          month = as.integer(month),
          Day = as.integer(Day)
        )
  
      all_data[[i]] <- res
      i <- i + 1
      }, silent = TRUE)
    }
  }
}
```

## 2.3 Merging data and creating variables

```{r}
#| code-fold: true
#| code-summary: "Click to view code for combining extracted data and creating date column"
climate_data <- bind_rows(all_data)

climate_data <- climate_data %>%
  mutate(
    # Create 'date' column from 'year', 'month', and 'day'
    date = make_date(year, month, Day)
  )

# View the first few rows to check the date column
head(climate_data)
```

```{r}
#| code-fold: true
#| code-summary: "Click to view code for merging aspatial data with station coordinates"
climate_data_2020to2024 <- left_join(climate_data, stations_df, by = c("Station" = "StationName"))

# Write the combined data to a separate CSV file
write_csv(climate_data, "data/climate_data_2020to2024.csv")

# View the first few rows of the merged data
head(climate_data_2020to2024)
```

## 3 Exploratory Data Analysis

## 4 Confirmatory Data Analysis

## 5 Geospatial Analysis & UI Prototype
