---
title: "Hands-on Exercise 9"
author: "Leng Yi Xiu"
date: "Mar 16 2025"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  message: false
  freeze: true
---

## 1 Modelling, Visualising and Analysing Network Data with R

### 1.1 Learning Outcomes

-   Create graph object data frames, manipulate them using appropriate functions of *dplyr*, *lubridate*, and *tidygraph*,

-   Build network graph visualisation using appropriate functions of *ggraph*,

-   compute network geometrics using *tidygraph*,

-   Build advanced graph visualisation by incorporating the network geometrics, and

-   Build interactive network visualisation using *visNetwork* package.

### 1.2 Loading R packages & Importing Data

```{r}
pacman::p_load(igraph, tidygraph, ggraph, 
               visNetwork, lubridate, clock,
               tidyverse, graphlayouts)
```

```{r}
GAStech_nodes <- read_csv("data/GAStech_email_node.csv")
GAStech_edges <- read_csv("data/GAStech_email_edge-v2.csv")
```

```{r}
glimpse(GAStech_nodes)
glimpse(GAStech_edges)
```

### 1.3 Data Wrangling

#### 1.3.1 Wrangling Time

::: callout-warning
The output report of GAStech_edges above reveals that the *SentDate* is treated as “Character” data type instead of "Date" data type. This is an error! Before we continue, it is important for us to change the data type of *SentDate* field back to “Date” data type.
:::

```{r}
GAStech_edges <- GAStech_edges %>%
  mutate(SendDate = dmy(SentDate)) %>%
  mutate(Weekday = wday(SentDate,
                        label = TRUE,
                        abbr = FALSE))
```

```{r}
glimpse(GAStech_edges)
```

#### 1.3.2 Data Aggregation

```{r}
GAStech_edges_aggregated <- GAStech_edges %>%
  filter(MainSubject == "Work related") %>%
  group_by(source, target, Weekday) %>%
    summarise(Weight = n()) %>%
  filter(source!=target) %>%
  filter(Weight > 1) %>%
  ungroup()
```

```{r}
glimpse(GAStech_edges_aggregated)
```

### 1.4 Creating network objects using tidygraph

In this section, we learn how to create a graph data model by using **tidygraph** package. It provides a tidy API for graph/network manipulation. While network data itself is not tidy, it can be envisioned as two tidy tables, one for node data and one for edge data. tidygraph provides a way to switch between the two tables and provides dplyr verbs for manipulating them. Furthermore it provides access to a lot of graph algorithms with return values that facilitate their use in a tidy workflow.

#### 1.4.1 The tbl_graph object

```{r}
GAStech_graph <- tbl_graph(nodes = GAStech_nodes,
                           edges = GAStech_edges_aggregated, 
                           directed = TRUE)
```

#### 1.4.2 Reviewing the output tidygraph's graph object

```{r}
GAStech_graph
```

::: callout-note
-   The output above reveals that *GAStech_graph* is a tbl_graph object with 54 nodes and 4541 edges.

-   The command also prints the first six rows of “Node Data” and the first three of “Edge Data”.

-   It states that the Node Data is **active**. The notion of an active tibble within a tbl_graph object makes it possible to manipulate the data in one tibble at a time.
:::

#### 1.4.3 Changing the active object

```{r}
GAStech_graph %>%
  activate(edges) %>%
  arrange(desc(Weight))
```

### 1.5 Plotting Static Network Graphs with ggraph package

As in all network graph, there are three main aspects to a **ggraph**’s network graph, they are:

-   [nodes](https://cran.r-project.org/web/packages/ggraph/vignettes/Nodes.html),

-   [edges](https://cran.r-project.org/web/packages/ggraph/vignettes/Edges.html) and

-   [layouts](https://cran.r-project.org/web/packages/ggraph/vignettes/Layouts.html).

#### 1.5.1 Plotting a basic network graph

```{r}
ggraph(GAStech_graph) +
  geom_edge_link() +
  geom_node_point()
```

::: callout-note
-   The basic plotting function is `ggraph()`, which takes the data to be used for the graph and the type of layout desired. Both of the arguments for `ggraph()` are built around *igraph*. Therefore, `ggraph()` can use either an *igraph* object or a *tbl_graph* object.
:::

#### 1.5.2 Changing the default network graph theme

```{r}
g <- ggraph(GAStech_graph) + 
  geom_edge_link(aes()) +
  geom_node_point(aes())

g + theme_graph()
```

#### 1.5.3 Changing the coloring of the plot

```{r}
g <- ggraph(GAStech_graph) + 
  geom_edge_link(aes(colour = 'grey50')) +
  geom_node_point(aes(colour = 'grey40'))

g + theme_graph(background = 'grey10',
                text_colour = 'white')
```

#### 1.5.4 Working with ggraph's layouts

📌 Fruchterman and Reingold layout

```{r}
g <- ggraph(GAStech_graph, 
            layout = "fr") +
  geom_edge_link(aes()) +
  geom_node_point(aes())

g + theme_graph()
```

📌circle layout

```{r}
g <- ggraph(GAStech_graph, 
            layout = "circle") +
  geom_edge_link(aes()) +
  geom_node_point(aes())

g + theme_graph()
```

#### 1.5.5 Modifying network nodes

```{r}
g <- ggraph(GAStech_graph, 
            layout = "nicely") + 
  geom_edge_link(aes()) +
  geom_node_point(aes(colour = Department, 
                      size = 3))

g + theme_graph()
```

#### 1.5.6 Modifying edges

```{r}
g <- ggraph(GAStech_graph, 
            layout = "nicely") +
  geom_edge_link(aes(width=Weight), 
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department), 
                  size = 3)

g + theme_graph()
```

### 1.6 Creating facet graphs

There are three functions in ggraph to implement faceting, they are:

-   [*facet_nodes()*](https://r4va.netlify.app/chap27) whereby edges are only draw in a panel if both terminal nodes are present here,

-   [*facet_edges()*](https://ggraph.data-imaginist.com/reference/facet_edges.html) whereby nodes are always drawn in al panels even if the node data contains an attribute named the same as the one used for the edge facetting, and

-   [*facet_graph()*](https://ggraph.data-imaginist.com/reference/facet_graph.html) faceting on two variables simultaneously.

#### 1.6.1 Working with `facet_edges()`

```{r}
set_graph_style()

g <- ggraph(GAStech_graph, 
            layout = "nicely") + 
  geom_edge_link(aes(width=Weight), 
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department), 
                  size = 2)

g + facet_edges(~Weekday)
```

The code chunk below uses *theme()* to change the position of the legend:

```{r}
set_graph_style()

g <- ggraph(GAStech_graph, 
            layout = "nicely") + 
  geom_edge_link(aes(width=Weight), 
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department), 
                  size = 2) +
  theme(legend.position = 'bottom')
  
g + facet_edges(~Weekday)
```

#### 1.6.2 A framed facet graph

```{r}
set_graph_style() 

g <- ggraph(GAStech_graph, 
            layout = "nicely") + 
  geom_edge_link(aes(width=Weight), 
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department), 
                  size = 2)
  
g + facet_edges(~Weekday) +
  th_foreground(foreground = "grey80",  
                border = TRUE) +
  theme(legend.position = 'bottom')
```

#### 1.6.2 Working with `facet_nodes()`

```{r}
set_graph_style()

g <- ggraph(GAStech_graph, 
            layout = "nicely") + 
  geom_edge_link(aes(width=Weight), 
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department), 
                  size = 2)
  
g + facet_nodes(~Department)+
  th_foreground(foreground = "grey80",  
                border = TRUE) +
  theme(legend.position = 'bottom')

```

### 1.7 Network Metrics Analysis

#### 1.7.1 Computing centrality indices

```{r}
g <- GAStech_graph %>%
  mutate(betweenness_centrality = centrality_betweenness()) %>%
  ggraph(layout = "fr") + 
  geom_edge_link(aes(width=Weight), 
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department,
            size=betweenness_centrality))
g + theme_graph()
```

::: callout-note
Things to learn from the code chunk above:

-   *mutate()* of **dplyr** is used to perform the computation.

-   the algorithm used, on the other hand, is the *centrality_betweenness()* of **tidygraph**.
:::

#### 1.7.2 Visualising network metrics

```{r}
g <- GAStech_graph %>%
  ggraph(layout = "fr") + 
  geom_edge_link(aes(width=Weight), 
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department, 
                      size = centrality_betweenness()))
g + theme_graph()
```

#### 1.7.3 Visualising Community

```{r}
g <- GAStech_graph %>%
  mutate(community = as.factor(group_edge_betweenness(weights = Weight, directed = TRUE))) %>%
  ggraph(layout = "fr") + 
  geom_edge_link(aes(width=Weight), 
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = community))  

g + theme_graph()
```

### 1.8 Building Interactive Network Graph with visNetwork

-   *visNetwork()* function uses a nodes list and edges list to create an interactive graph.

    -   The nodes list must include an “id” column, and the edge list must have “from” and “to” columns.

    -   The function also plots the labels for the nodes, using the names of the actors from the “label” column in the node list.

-   The resulting graph is fun to play around with.

    -   You can move the nodes and the graph will use an algorithm to keep the nodes properly spaced.

    -   You can also zoom in and out on the plot and move it around to re-center it.

#### 1.8.1 Data Preparation

```{r}
GAStech_edges_aggregated <- GAStech_edges %>%
  left_join(GAStech_nodes, by = c("sourceLabel" = "label")) %>%
  rename(from = id) %>%
  left_join(GAStech_nodes, by = c("targetLabel" = "label")) %>%
  rename(to = id) %>%
  filter(MainSubject == "Work related") %>%
  group_by(from, to) %>%
    summarise(weight = n()) %>%
  filter(from!=to) %>%
  filter(weight > 1) %>%
  ungroup()
```

#### 1.8.2 Plotting the first interactive network graph

```{r}
visNetwork(GAStech_nodes, 
           GAStech_edges_aggregated)
```

#### 1.8.3 Working with layout

```{r}
visNetwork(GAStech_nodes,
           GAStech_edges_aggregated) %>%
  visIgraphLayout(layout = "layout_with_fr") 
```

#### 1.8.4 Working with visual attributes - Nodes

```{r}
GAStech_nodes <- GAStech_nodes %>%
  rename(group = Department) 
```

```{r}
visNetwork(GAStech_nodes,
           GAStech_edges_aggregated) %>%
  visIgraphLayout(layout = "layout_with_fr") %>%
  visLegend() %>%
  visLayout(randomSeed = 123)
```

#### 1.8.5 Working with visual attributes - Edges

```{r}
visNetwork(GAStech_nodes,
           GAStech_edges_aggregated) %>%
  visIgraphLayout(layout = "layout_with_fr") %>%
  visEdges(arrows = "to", 
           smooth = list(enabled = TRUE, 
                         type = "curvedCW")) %>%
  visLegend() %>%
  visLayout(randomSeed = 123)
```

#### 1.8.6 Interactivity

```{r}
visNetwork(GAStech_nodes,
           GAStech_edges_aggregated) %>%
  visIgraphLayout(layout = "layout_with_fr") %>%
  visOptions(highlightNearest = TRUE,
             nodesIdSelection = TRUE) %>%
  visLegend() %>%
  visLayout(randomSeed = 123)
```
